Divergent elliptical biface morphology: Distinct local retouch practice or Serpent Dance proxy for elapids and vipers?

### 2022-12-24 supp/index.qmd
<edit index>

---
title: "Divergent elliptical biface morphology: Distinct local retouch practice or serpent dance proxy for Elapids and Vipers?"
author:
  - name: Robert Z. Selden, Jr.
    affiliations:
      - Heritage Research Center, Stephen F. Austin State University; Department of Biology, Stephen F. Austin State University; Texas Archeological Research Laboratory, The University of Texas at Austin; Cultural Heritage Department, Jean Monnet University
    email: zselden@sfasu.edu
    orcid: 0000-0002-1789-8449
  - name: John E. Dockall
    affiliations:
      - Stantec, Inc.
    email: john.dockall@stantec.com
    orcid: 0000-0002-0940-7144
  - name: David H. Dye
    affiliations: 
      - Department of Earth Sciences, The University of Memphis
    email: daviddye@memphis.edu
bibliography: [book.bib]
date: "`r format(Sys.time(), '%d %B, %Y')`"
link-citations: yes
github-repo: seldenlab/elliptical.bifaces
---

@@

### Keywords {-}

American Southeast, Caddo, NAGPRA, computational archaeology, archaeoinformatics, museum studies, digital humanities, non-Western art history, geometric morphometrics, STEM, STEAM, ovoid biface, Galt biface, Jowell knife, Jowell knives, Jowell biface, Nikolas knife, lithics, parallel-oblique flaking

@@

1.   Elliptical bifaces found at Millsap Cache were manufactured using Kay County Chert [@RN11461], while those from Jowell Farm were manufactured using Edwards Chert. Differences in stone tool shape as a function of raw material have been reported for Perdiz arrow points in the region [@RN9364]; however, others have demonstrated that the attribution of morphological variability to raw material differences is an assumption that warrants regular testing, suggesting that primacy should not be assumed [@RN8442]. Thus, a comparison of between-site morphology was included to assess elliptical biface morphology as a function of raw material.
  a)    The biogeography of elapid and viper species differs within the ancestral Caddo area, where elapid populations occur south of the Red River, excluding them from the geography surrounding Millsap Cache. If elliptical biface morphology was intended to represent local elapid and viper populations, then elliptical bifaces from these sites will differ due to differences in the constituent communities of elapids and vipers.

2.    The elliptical bifaces from Jowell Farm were discovered in an area of the ancestral Caddo region where others have assumed large bifaces to be products of external manufacture due to the absence of local production evidence and the availability of raw material/s of similar size and quality [@RN3682;@RN11783;@RN21001]. Given the potential that elliptical bifaces were produced by knappers outside of the ancestral Caddo region for at least one of the caches (Jowell Farm), it is assumed that the smaller elliptical bifaces will more readily reflect morphology imposed by Caddo knappers through resharpening and/or retouch incurred through normal use [@RN11264]. Thus, it is expected that elliptical bifaces will express convergent morphology, where morphological consistency is achieved through the idiosyncratic approaches to resharpening and/or retouch by Caddo knappers in the small size class.
  a)   Elapid and viper species will progress through the smaller size class during normal developmental stages before reaching adulthood, where their morphology differs within and among species. If elliptical biface morphology is thought to represent distinct elapid and viper populations, then the most pronounced morphological difference will occur in the large size class among adults. Thus, if elliptical bifaces were intended to represent elapid and viper species, then the morphological pattern will be identical to H2. However, analytical results would be read in reverse, where expected similarities occur at the smaller size class, before diverging in the large size class.

3.    A measure of morphological disparity was subsequently used to assess whether biface morphology may have been more morphologically diverse in the large size class, suggestive of greater diversity in biface shape or size related to production activities [@RN11264]. An alternative hypothesis is that the idiosyncratic responses to resharpening and/or retouch by Caddo knappers required to address damage incurred through normal use generate greater morphological diversity in the small size class [@RN11264].
  a)    Elapid and viper species express greater morphological diversity at their larger size. Thus, if elliptical bifaces were intended to represent elapid and viper species, and biological diversity differs between sites---elapid biogeography restricted to the area of Jowell Farm---then morphological disparity will occur between sites in the large size class.

>>>>

---
title: "Divergent elliptical biface morphology: Distinct local retouch practice or Serpent Dance proxy for elapids and vipers?"
author:
  - name: Robert Z. Selden, Jr.
    affiliations:
      - Heritage Research Center, Stephen F. Austin State University; Department of Biology, Stephen F. Austin State University; Texas Archeological Research Laboratory, The University of Texas at Austin; Cultural Heritage Department, Jean Monnet University
    email: zselden@sfasu.edu
    orcid: 0000-0002-1789-8449
  - name: John E. Dockall
    affiliations:
      - Stantec, Inc.
    email: john.dockall@stantec.com
    orcid: 0000-0002-0940-7144
  - name: David H. Dye
    affiliations: 
      - Department of Earth Sciences, The University of Memphis
    email: daviddye@memphis.edu
bibliography: [book.bib]
date: "`r format(Sys.time(), '%d %B, %Y')`"
link-citations: yes
github-repo: seldenlab/elliptical.bifaces
description: Supplementary Materials
---

@@

### Keywords

_American Southeast, Caddo, NAGPRA, computational archaeology, archaeoinformatics, museum studies, digital humanities, non-Western art history, geometric morphometrics, STEM, STEAM, ovoid biface, Galt biface, Jowell knife, Jowell knives, Jowell biface, Nikolas knife, lithics, parallel-oblique flaking_

@@

1.   Elliptical bifaces found at Millsap Cache were manufactured using Kay County Chert [@RN11461], while those from Jowell Farm were manufactured using Edwards Chert. Differences in stone tool shape as a function of raw material have been reported for Perdiz arrow points in the region [@RN9364]; however, others have demonstrated that the attribution of morphological variability to raw material differences is an assumption that warrants regular testing, suggesting that primacy should not be assumed [@RN8442]. Thus, a comparison of between-site morphology was included to assess elliptical biface morphology as a function of raw material.
  a)    The biogeography of elapid and viper species differs within the ancestral Caddo area, where elapids are absent from the geography surrounding Millsap Cache, and one viper species (_Crotalus atrox_, or western/Texas diamondback rattlesnake) is absent from the geography surrounding Jowell Farm. If elliptical biface morphology was intended to represent local elapid and viper populations, then elliptical bifaces from these sites will differ due to differences in the constituent communities of elapids and vipers.

2.    The elliptical bifaces from Jowell Farm were discovered in an area of the ancestral Caddo region where others have assumed large bifaces to be products of external manufacture due to the absence of local production evidence and the availability of raw material/s of similar size and quality [@RN3682;@RN11783;@RN21001]. Given the potential that elliptical bifaces were produced by knappers outside of the ancestral Caddo region for at least one of the caches (Jowell Farm), it is assumed that the smaller elliptical bifaces will more readily reflect morphology imposed by Caddo knappers through resharpening and/or retouch incurred through normal use [@RN11264]. Thus, it is expected that elliptical bifaces will express convergent morphology, where morphological consistency is achieved through the idiosyncratic approaches to resharpening and/or retouch by Caddo knappers in the _`small size class`_.
  a)   Elapid and viper species progress through the smaller size class during normal developmental stages before reaching adulthood, where their morphology differs within and among species. If elliptical biface morphology is thought to represent distinct elapid and viper populations, then the most pronounced morphological difference will occur in the _`large size class`_, associated with adult snakes. Thus, if elliptical bifaces were intended to represent elapid and viper species, then the morphological pattern will be identical to H2. However, analytical results would be read in reverse, where expected similarities occur at the smaller size class, before diverging in the _`large size class`_.

3.    A measure of morphological disparity was subsequently used to assess whether biface morphology may have been more morphologically diverse in the _`large size class`_, suggestive of greater diversity in biface shape or size related to production activities [@RN11264]. An alternative hypothesis is that the idiosyncratic responses to resharpening and/or retouch by Caddo knappers required to address damage incurred through normal use generate greater morphological diversity in the _`small size class`_ [@RN11264].
  a)    Elapid and viper species express greater morphological diversity at their larger size. Thus, if elliptical bifaces were intended to represent elapid and viper species, and biological diversity differs between sites---elapid biogeography restricted to the area of Jowell Farm---then morphological disparity will occur between sites in the _`large size class`_.
